Reasoning Graphically about Quantum Computation

### 2008-03-07 papers/reasoning-graphically/paper.tex
work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1405 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\begin{definition}
  \emph{copy}: copies a !-box, $b$ in a graph to produce a new graph
  with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
  is the new one. The set of !-boxes in the copied graph now also
  contains the new !-box $b'$. Any edges between a node, $n$, inside
  the !-box $b$, and a node, $m$, outside it, get copied so that there
  is a new edge from $m$ to the new copy of $n$ in $b'$.
\end{definition}

\begin{definition}
  \emph{drop} simply removes the !-box, but leaves its contents in the
  graph.
\end{definition}

\begin{definition}
  \emph{kill} removes from the graph all nodes in the !-box as well as
  any incident edges.
\end{definition}

\begin{definition}
  \emph{merge} combines two !-boxes, $B_1$ and $B_2$ into a single
  larger !-box $B_1 \cup B_2$.
\end{definition}

@@

As well as these equations, recall that the object level formalism
must define the data held at the nodes and edges as well as provide a
suitable macthing algorithm for these. Our model of quantum
computation provides no data for the edges. The nodes, on the other
hand, are either a $H$ with no additional data (a Hadamard gate), or
have an \emph{angle} and are either the $X$ or $Z$ transforms. For
their part, angles are expressed as rational numbers which correspond
to the coefficient of $\pi$ in the corresponding matrix.

The result is that we give the $X$ and $Z$ nodes an \emph{angle
  expression}. When a node is within a !-box, this is a variable which
gets instantiated to a new variable in each of the unfoldings of the
!-box. When a node is not within a !-box, angle-expressions are a
mapping from a set of angle-variables to the corresponding rational
cooefficient. When an angle-expression contains a variable within a
!-box, this is interpreted as a sum of the variables that result from
it's unfolding. This rather simple expression language has a simple
normal form by lexographically ordering the angle-variable. Matching
then results in variables being instantiated and all affected nodes
normalising the result of the instantiation. The substitution rule
also has to make sure that a angle-variables in the rule being applied
are distinct from those in the expression being rewritten.

@@

\subsection{Computing the Quantum Fourier Transform}
\label{ssec:cs:quantum-transform}

The quantum fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}.

\begin{figure}[t]
$$\begin{array}{cccc}
  & \inlinegraphic{2cm}{qft2} & = & \inlinegraphic{2cm}{qft3} \\ 
 & & & \\
= & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10}
\end{array}$$
\label{fig:quantum-transform}\caption{An example computation of the
  Quantum Fourier Transform, performed symbollically by rewriting.}
\end{figure}

>>>>

\begin{description}
  \item[copy]: copies a !-box, $b$ in a graph to produce a new graph
  with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
  is the new one. The set of !-boxes in the copied graph now also
  contains the new !-box $b'$. Any edges between a node, $n$, inside
  the !-box $b$, and a node, $m$, outside it, get copied so that there
  is a new edge from $m$ to the new copy of $n$ in $b'$.

\item[drop]: simply removes the !-box, but leaves its contents in the
  graph.

\item[kill]: removes from the graph all nodes in the !-box as well as
  any incident edges.

\item[merge]: combines two !-boxes, $B_1$ and $B_2$ into a single
  larger !-box $B_1 \cup B_2$.
\end{description}

@@

Our model of quantum computation requires no data for the edges. The
nodes, on the other hand, are either a $H$ with no additional data (a
Hadamard gate), or have an \emph{angle} and are either the $X$ or $Z$
transforms. For their part, angles are expressed as rational numbers
which correspond to the coefficient of $\pi$ in the corresponding
matrix.

The result is that we give the $X$ and $Z$ nodes an \emph{angle
  expression}. When a node is within a !-box, this is a variable which
gets instantiated to a new variable in each of the unfoldings of the
!-box. When a node is not within a !-box, angle-expressions are a
mapping from a set of angle-variables to the corresponding rational
cooefficient. When an angle-expression contains a variable within a
!-box, this is interpreted as a sum of the variables that result from
it's unfolding. This rather simple expression language has a simple
normal form by lexographically ordering the angle-variable. Matching
then results in variables being instantiated and the values of all
affected nodes are then normalised. An additional implementation
detail must also be observed for the substitution rule: it must ensure
that angle-variables in the rule being applied are distinct from those
in the expression being rewritten.

@@

\subsubsection{Computing the Quantum Fourier Transform.}
\label{ssec:cs:quantum-transform}

The quantum fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
polynomial time algorithm for factoring. In our graph patten calculus
this ciruit becomes the top-left graph in
Figure~\ref{fig:quantum-transform}. This figure shows how computation
can be symbolically performed by rewriting with the lifted equations
from Figure~\ref{fig:graph-quant-eqns} with the graph pattern version
of the Spider Theorem.

\begin{figure}[t]
\begin{tabular}{cccc}
\hline
 & & & \\
  & \inlinegraphic{2cm}{qft2} & = & \inlinegraphic{2cm}{qft3} \\ 
 & & & \\
\hline
 & & & \\
= & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10} \\
 & & & \\
\hline
\end{tabular}
\label{fig:quantum-transform}\caption{An example computation of the
  Quantum Fourier Transform with inputs 1 and 0, performed
  symbollically by rewriting. }
\end{figure}
